Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms

### 2018-04-16 paper/paper.tex
IJCAI doesn't love us. ESA?

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in
\usepackage{ijcai18}
\usepackage{times}
\usepackage{xcolor}
\usepackage{soul}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{tikz}

\usepackage{cleveref}

@@

@@

\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}
\newcommand{\citep}[1]{\cite{#1}}
\newcommand{\citepos}[1]{\citeauthor{#1}'s \shortcite{#1}}

@@

@@

% cref style
\crefname{algorithm}{Algorithm}{Algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}

\title{Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This footnote
leaves room for a piece of text saying how grateful we are to our funding agency. This includes a
magic grant number, so it takes up three precious lines of space.}}
\author{Anonymous \and Anonymous \and Anonymous \\ Somewhere \\ Some Email Address}

@@

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of these changes are shown in
\cref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\cref{figure:scatter-dds}) or by random value-ordering with restarts
(\cref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing benefits in two maximum common induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. The
\emph{neighbourhood degree sequence} of $v$, $\nds_G(v)$, is the sequence of degrees of the
neighbours of $v$ in descending order, and we denote by $\preceq$ the inclusion relation of
\citet{DBLP:journals/constraints/ZampelliDS10}. When the graph is clear from the context, we omit
the $G$ subscripts. As per \citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length exactly $\ell$ between vertices $v$ and $w$ in $G$.

@@

We now briefly outline the non-shaded parts of \cref{algorithm:sip}, which is how we will solve the
subgraph isomorphism problem. This algorithm is very closely based upon the $k{\downarrow}$
algorithm of \citet{DBLP:conf/aaai/HoffmannMR17} with $k = 0$, and we refer the reader to that paper
for full technical details; that algorithm, in turn, is a simplification of the Glasgow2 algorithm
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.

\Cref{algorithm:sip} works by recursively building up a valid mapping of pattern vertices to target
vertices. For each pattern vertex $v$, we have a domain $D_v$ containing a set of feasible target
vertices. Each domain is initialised at the top of search (\siplinerangeref{initstart}{initend}),
using the invariants described by \citet{DBLP:conf/aaai/HoffmannMR17} to restrict its initial
values.

@@

Our $\FuncSty{propagate}$ routine is also close to that of \citet{DBLP:conf/aaai/HoffmannMR17},
differing only in that we carry out the filtering for injectivity
(\siplinerangeref{alldiffstart}{alldiffend}) after every assignment rather than once at the end of
the unit propagation loop. (We make this change because the all-different propagator used is not
idempotent, and we wish to avoid an odd interaction when we introduce nogood filtering.)

@@

        (\mathcal{P}^{2,2},\ \mathcal{T}^{2,2}), $\\\hspace*{6em}$ (\mathcal{P}^{3,2},\ \mathcal{T}^{3,2}),

@@

        \nl \ForEach{$(P,\,T) \in L$}{
            \nl $D_v \gets \{ w \in D_v : $ \siplabel{degreefiltering} \\\hspace*{4em}$v \sim_P v \Rightarrow w \sim_T w~\wedge$
            \\\hspace*{4em}$\nds_P(v) \preceq \nds_T(w)\}$\siplabel{initend} \;

@@

            \nl \ForEach{$(P,\,T) \in L$}{
                \nl \lIf{$v \sim_P w$}{$D_w \gets D_w \cap \neighbourhood_T(v')$

@@

    \nl $D_v \gets D_v \setminus H$ \siplabel{elimhall} \;
    \nl $(A,\,n) \gets (A \cup D_v,\,n + 1)$ \siplabel{acc} \;

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we use this same dynamic ordering throughout the paper, so as not to
affect the performance of the algorithm on unsatisfiable instances.  For value ordering, which
occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will initially select vertices from
highest degree to lowest degree; we return to value ordering in the second half of this paper. Note
that the variable-ordering heuristic affects both unsatisfiable and satisfiable instances, whilst
the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We implemented\footnote{Institutional DOI URL removed for anonymous review}
\cref{algorithm:sip} in C++ using the bit-parallel implementation of
\citet{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

We use the dataset introduced by \citet{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers.
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least \input{gen-n-sat.tex}\unskip\ of these instances are known to be satisfiable, and at least
\input{gen-n-unsat.tex}\unskip\ are unsatisfiable.

@@

We begin by verifying that our implementation of the basic \cref{algorithm:sip} is competitive. In
\cref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \cref{algorithm:sip} is derived
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \citep{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\citep{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout; the horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

As expected, the performance of \cref{algorithm:sip} with the degree heuristic is very close to that
of Glasgow2, although we perform better at lower runtimes because we use simpler preprocessing at
the top of search. Our implementation also clearly beats PathLAD and VF2, except for very low
choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the second
plot in \cref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.

@@

\Cref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable instances
solved over time using different value-ordering heuristics:

\paragraph{Degree} We select vertices from highest degree to lowest degree
\citep{DBLP:conf/ijcai/McCreeshPT16}.

\paragraph{Random} We select vertices uniformly at random.

\paragraph{Anti} We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: as expected, degree beats random, and random beats anti.

\paragraph{Biased} We branch by selecting a vertex $v'$ from the chosen domain $D_v$ with
probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in D_v}{2^{\deg(w)}}} \text{.} \] This heuristic
continues to prefer vertices of high degree, but will give the same chance of being selected next to
two vertices of equal degree.  It also introduces an element of randomness, which is needed to make
restarts have an effect.

The biased heuristic resembles the well-known \emph{softmax} weighting scheme applied to the degree
heuristic, although usually \emph{softmax} uses a base of $e$ rather than $2$.  Experiments showed
that in this context, the choice of base is not important to the search space size; however, a base
of $2$ avoids the need to use floating point arithmetic, which was a substantial overhead.

The performance of the degree and biased heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Cref{figure:scatter-heuristics} gives a more detailed look at these results. This scatter plot
shows that despite the aggregate performance being very similar, on a case by case basis, the two
heuristics can make a large difference to the performance of satisfiable instances.

@@

A \emph{discrepancy} is where search goes against a value-ordering heuristic.  Discrepancy searches
\citep{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97} are alternatives to
backtracking search that initially search disallowing all discrepancies, and then retry search
allowing an increasing number of discrepancies at each iteration until either a solution is found or
unsatisfiability is proven. These schemes assume that value-ordering heuristics are usually
reliable, and that most solutions can be found with only a small number of discrepancies; in such
cases, the heavy commitment to early branching choices made by conventional backtracking search can
be extremely costly.

\Cref{figure:scatter-dds} shows the effects of adding \citepos{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\citep{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon, although \citet{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering
heuristics from restarts, and \citet{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of
parallel search.

The shaded code shows how to introduce restarts to \cref{algorithm:sip}. The recursive
$\FuncSty{search}$ procedure is modified to return either true (if the instance is satisfiable),
false (unsatisfiable), or a third value if some threshold was reached without the solution being
known.  The main function then calls $\FuncSty{search}$ in a loop (\siplineref{restartsloop}), until
a solution is found.

We use the Luby scheme \citep{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used SMAC \citep{DBLP:conf/lion/HutterHL11} to select the value 660. (We also tried a geometric
restart scheme, which gave less favourable results.)

@@

Finally, we use the two watched literals technique \citep{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods (\siplineref{2wl}). This has two benefits: the propagation complexity does
not particularly depend upon the number of stored nogoods, and it does not require any work upon
backtracking.  Other more sophisticated nogood generation and propagation schemes exist
\citep{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}, but it is not clear whether these will be
beneficial in a setting where we are not maintaining arc consistency.

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Cref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts.

@@

\Cref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.

In the more detailed view in \cref{figure:scatter-by-family}, comparing the basic algorithm with the
degree heuristic to the degree-biased algorithm with restarts, all of the unsatisfiable instances
are very close to the $x-y$ diagonal, showing that their performance is nearly unchanged. On the
other hand, there are large numbers of satisfiable instances well below the diagonal line,
indicating large speedups.  Better yet, there are only a handful of satisfiable instances that are
more than a factor of ten times worse. In other words, as well as improving performance, we have
made up most of the consistency we lost by introducing randomness.

Having looked at subgraph isomorphism in detail, we now briefly discuss the more general maximum
common induced subgraph problem.  Two recent algorithms for this problem also make use of
backtracking search with degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\citep{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only small changes
to handle wildcards in nogoods.

Meanwhile, the McSplit${\downarrow}$ algorithm
\citep{DBLP:conf/ijcai/McCreeshPT17} uses a constraint programming style search, but with special
propagators and backtrackable data structures that exploit special properties of the problem. The
unconventional domain store used by McSplit${\downarrow}$ precludes the use of arbitrary unit
propagation, and so when introducing restarts, we cannot propagate using nogoods.  Instead, we can
only detect when we are inside an already-visited branch.  We must therefore use the one watched
literal scheme instead, and we also introduce a basic subsumption scheme to prune redundant clauses.

Performance results from these two modified algorithms, using the same families of instances as in
the previous section, are shown in \cref{figure:kdown,figure:mcsplit,figure:mcs}. Although we have
moved from a decision problem to an optimisation problem, the same changes remain clearly
beneficial. For the $k{\downarrow}$ algorithm, the change has a minimal effect on many instances
(typically, where the $k = 0$ subproblem is unsatisfiable and hard, and the $k = 1$ subproblem is
satisfiable and easy), but gives large benefits on many more instances than it penalises: it is over
an order of magnitude better on nearly a hundred instances, whilst being an order of magnitude worse
on only seven.

@@

>>>>

\documentclass[a4paper,UKenglish]{lipics-v2018}

\usepackage{microtype}
\usepackage{complexity}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{tikz}

@@

\bibliographystyle{plainurl}

@@

@@

\newcommand{\figureref}[1]{Figure~\ref{#1}}
\newcommand{\Figureref}[1]{Figure~\ref{#1}}
\newcommand{\threefigureref}[3]{Figures~\ref{#1}, \ref{#2} and \ref{#3}}
\newcommand{\algorithmref}[1]{Algorithm~\ref{#1}}
\newcommand{\Algorithmref}[1]{Algorithm~\ref{#1}}

@@

\title{Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms}
\author{Ciaran McCreesh}{University of Glasgow}{ciaran.mccreesh@glasgow.ac.uk}{}{}
\author{Patrick Prosser}{University of Glasgow}{patrick.prosser@glasgow.ac.uk}{}{}
\author{James Trimble}{University of Glasgow}{j.trimble.1@research.gla.ac.uk}{}{}

\authorrunning{C. McCreesh, P. Prosser and J. Trimble}

\Copyright{Ciaran McCreesh, Patrick Prosser and James Trimble}

\subjclass{
\ccsdesc[500]{Mathematics of computing~Graph algorithms},
\ccsdesc[500]{Theory of computation~Discrete optimization},
\ccsdesc[500]{Theory of computation~Backtracking}}

\keywords{Subgraph isomorphism, combinatorial search, rich search}

\funding{This work was supported by the Engineering and Physical Sciences Research Council [grant
numbers EP/026842/1 and EP/M508056/1]}

\acknowledgements{The authors would like to thank Ruth Hoffmann, Christophe Lecoutre, Christine
Solnon and Craig Reilly for their comments.}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{26th Annual European Symposium on Algorithms (ESA 2018)}
\EventShortTitle{ESA 2018}
\EventAcronym{ESA}
\EventYear{2018}
\EventDate{August 20--22, 2018}
\EventLocation{Helsinki, Finland}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%\nolinenumbers %uncomment to disable line numbering
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of these changes are shown in
\figureref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\figureref{figure:scatter-dds}) or by random value-ordering with restarts
(\figureref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing benefits in two maximum common induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. The
\emph{neighbourhood degree sequence} of $v$, $\nds_G(v)$, is the sequence of degrees of the
neighbours of $v$ in descending order, and we denote by $\preceq$ the inclusion relation of
Zampelli et al.\ \cite{DBLP:journals/constraints/ZampelliDS10}. When the graph is clear from the context, we omit
the $G$ subscripts. As per Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length exactly $\ell$ between vertices $v$ and $w$ in $G$.

@@

We now briefly outline the non-shaded parts of \algorithmref{algorithm:sip}, which is how we will solve the
subgraph isomorphism problem. This algorithm is very closely based upon the $k{\downarrow}$
algorithm of Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} with $k = 0$, and we refer the reader to that paper
for full technical details; that algorithm, in turn, is a simplification of the Glasgow2 algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.

\Algorithmref{algorithm:sip} works by recursively building up a valid mapping of pattern vertices to target
vertices. For each pattern vertex $v$, we have a domain $D_v$ containing a set of feasible target
vertices. Each domain is initialised at the top of search (\siplinerangeref{initstart}{initend}),
using the invariants described by Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} to restrict its initial
values.

@@

Our $\FuncSty{propagate}$ routine is also close to that of Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17},
differing only in that we carry out the filtering for injectivity
(\siplinerangeref{alldiffstart}{alldiffend}) after every assignment rather than once at the end of
the unit propagation loop. (We make this change because the all-different propagator used is not
idempotent, and we wish to avoid an odd interaction when we introduce nogood filtering.)

@@

        (\mathcal{P}^{2,2},\ \mathcal{T}^{2,2}), (\mathcal{P}^{3,2},\ \mathcal{T}^{3,2}),

@@

        \nl \lForEach{$(P,\,T) \in L$}{$D_v \gets \{ w \in D_v : $ \siplabel{degreefiltering} $v \sim_P v \Rightarrow w \sim_T w~\wedge$
            $\nds_P(v) \preceq \nds_T(w)\}$\siplabel{initend}

@@

            \nl \lForEach{$(P,\,T) \in L~\KwSty{where}~v \sim_P w$}{$D_w \gets D_w \cap \neighbourhood_T(v')$}

@@

    \nl $(D_v, A,\,n) \gets (D_v \gets D_v \setminus H\siplabel{elimhall}, A \cup D_v,\,n + 1)$ \siplabel{acc} \;

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \algorithmref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we use this same dynamic ordering throughout the paper, so as not to
affect the performance of the algorithm on unsatisfiable instances.  For value ordering, which
occurs on \siplineref{valueordering} of \algorithmref{algorithm:sip}, we will initially select vertices from
highest degree to lowest degree; we return to value ordering in the second half of this paper. Note
that the variable-ordering heuristic affects both unsatisfiable and satisfiable instances, whilst
the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{\url{https://github.com/ciaranm/ijcai2018-restarting-sip-paper}, DOI to be issued pre-publication}
\algorithmref{algorithm:sip} in C++ using the bit-parallel implementation of
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

We use the dataset introduced by Kotthoff et al.\ \cite{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers.
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least \input{gen-n-sat.tex}\unskip\ of these instances are known to be satisfiable, and at least
\input{gen-n-unsat.tex}\unskip\ are unsatisfiable.

@@

We begin by verifying that our implementation of the basic \algorithmref{algorithm:sip} is competitive. In
\figureref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \algorithmref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\cite{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout; the horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

As expected, the performance of \algorithmref{algorithm:sip} with the degree heuristic is very close to that
of Glasgow2, although we perform better at lower runtimes because we use simpler preprocessing at
the top of search. Our implementation also clearly beats PathLAD and VF2, except for very low
choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the second
plot in \figureref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.

@@

\Figureref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable instances
solved over time using different value-ordering heuristics:

\begin{description}
\item[Degree] We select vertices from highest degree to lowest degree
\cite{DBLP:conf/ijcai/McCreeshPT16}.

\item[Random] We select vertices uniformly at random.

\item[Anti] We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: as expected, degree beats random, and random beats anti.

\item[Biased] We branch by selecting a vertex $v'$ from the chosen domain $D_v$ with
probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in D_v}{2^{\deg(w)}}} \text{.} \]\end{description}

The biased heuristic continues to prefer vertices of high degree, but will give the same chance of
being selected next to two vertices of equal degree.  It also introduces an element of randomness,
which is needed to make restarts have an effect.  We note that it resembles the well-known
\emph{softmax} weighting scheme applied to the degree heuristic, although usually \emph{softmax}
uses a base of $e$ rather than $2$.  Experiments showed that in this context, the choice of base is
not important to the search space size; however, a base of $2$ avoids the need to use floating point
arithmetic, which was a substantial overhead.

The performance of the degree and biased heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Figureref{figure:scatter-heuristics} gives a more detailed look at these results. This scatter plot
shows that despite the aggregate performance being very similar, on a case by case basis, the two
heuristics can make a large difference to the performance of satisfiable instances.

@@

A \emph{discrepancy} is where search goes against a value-ordering heuristic.  Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97} are alternatives to
backtracking search that initially search disallowing all discrepancies, and then retry search
allowing an increasing number of discrepancies at each iteration until either a solution is found or
unsatisfiability is proven. These schemes assume that value-ordering heuristics are usually
reliable, and that most solutions can be found with only a small number of discrepancies; in such
cases, the heavy commitment to early branching choices made by conventional backtracking search can
be extremely costly.

\Figureref{figure:scatter-dds} shows the effects of adding Walsh's {DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \algorithmref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \algorithmref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon, although Razgon et al.\ \cite{DBLP:conf/flairs/RazgonOP07} look at learning value-ordering
heuristics from restarts, and Chu et al.\ \cite{DBLP:conf/cp/ChuSS09} use a similar scheme in the context of
parallel search.

The shaded code shows how to introduce restarts to \algorithmref{algorithm:sip}. The recursive
$\FuncSty{search}$ procedure is modified to return either true (if the instance is satisfiable),
false (unsatisfiable), or a third value if some threshold was reached without the solution being
known.  The main function then calls $\FuncSty{search}$ in a loop (\siplineref{restartsloop}), until
a solution is found.

We use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used SMAC \cite{DBLP:conf/lion/HutterHL11} to select the value 660. (We also tried a geometric
restart scheme, which gave less favourable results.)

@@

Finally, we use the two watched literals technique \cite{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods (\siplineref{2wl}). This has two benefits: the propagation complexity does
not particularly depend upon the number of stored nogoods, and it does not require any work upon
backtracking.  Other more sophisticated nogood generation and propagation schemes exist
\cite{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}, but it is not clear whether these will be
beneficial in a setting where we are not maintaining arc consistency.

@@

In \figureref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random value-ordering heuristic comfortably beats the degree strategy without restarts. In other
words, although having a good value-ordering heuristic is beneficial, introducing randomness into
the search is better, if it is done alongside a mechanism to avoid heavy commitment to any
particular random choice. \Figureref{figure:scatter-random} looks at this in more detail. The performance
on unsatisfiable instances, drawn as dark dots, remains unchanged as expected (which shows that the
overheads of nogoods and restarting are negligible). However, although better in aggregate on
satisfiable instances, on a case by case basis many instances perform substantially worse with
randomisation and restarts.

@@

@@

\Figureref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.
